### 2013-11-11 src2/thesis.tex
update

ArBB is based on a set of parallel primitives (or structures) on dense one, two and three dimensional vectors and nested vectors: 
\begin{itemize} 
\item {\bf Reductions:}   {\tt add\_reduce}, {\tt mul\_reduce}, {\tt max\_reduce} ... 
\item {\bf Scans:}        {\tt add\_scan}, {\tt mul\_scan}, {\tt max\_scan} ... 
\item {\bf Sorting:}      {\tt sort} 
\item {\bf Permutations:} {\tt gather}, {\tt scatter}, {\tt reverse} ...
\item {\bf Sequential loops:} {\tt \_for}, {\tt \_while} 
\item {\bf Map, zipWith, stencil:}  {\tt map} 
\end{itemize}

These operations (and many more, there are about 120 operations in total) can be used 
by the programmer when writing programs using ArBB. Note that there are sequential looping 
constructs with special names, {\tt \_for} and {\tt \_while}. ArBB is a deeply embedded 
language and calling the ArBB functions does not actually execute them, it just ads a node to 
some internal representation. Therefore, the use of normal C++ for loops leads to unrolled 
ArBB programs and a special {\tt \_for} loop is needed to get a loop in the generated program. 
This is exactly the way these things work in Haskell embedded languages, but it may be even 
more surprising to a C++ programmer. Before an ArBB function can be computed it needs to be 
captured. For this, ArBB has a {\tt call} function that is used to run functions using ArBB 
functionality. The first time {\tt call} is used on a function pointer that function is 
executed and the ArBB functions used in it build an AST (we say the function has been 
captured). The AST is then compiled and specialised for the hardware available. Any 
subsequent calls of a function using ArBB functionality do not lead to a recompilation, 
the first call caches this compiled code.

@@

The ArBB Haskell bindings provide a very rudimentary interface to ArBB functionality and 
is not useful for application implementation. A higher-level interface to ArBB to offer 
the Haskell programmer is needed. Our first attempt was to implement an Accelerate backend 
using the ArBB bindings. However, this had some hard problems to solve. Accelerate is a 
more richer language and there was an API mismatch between Accelerate and ArBB. These problems 
are mentioned in in paper G, section~\ref{sec:paperG}. As a way to offer ArBB functionality 
to the Haskell in a way that is useful and simpler to implement, we start working on EmbArBB.

With EmbArBB we improve the interfacing with Haskell by implementing a deeply embedded language. 
Now, programs in EmbArBB create ASTs. These ASTs are compiled via the ArBB bindings creating 
function objects that can then be run.

@@

ArBB supports one, two and three dimensional arrays, called dense vectors. In EmbArBB we 
represent these with a data type, {\tt DVector}, parameterised by dimensionality and 
element type. The dimension parameter can be either {\tt Dim1}, {\tt Dim2} or {\tt Dim3}. 
There are also nested vectors, in EmbArBB called {\tt NVector}, parameterised on element type.

@@

EmbArBB provides many of the ArBB operations on vectors but provides a 
shape-polymorphic interface to them. For example, a reduction operation can take a two  
dimensional vector and reduce all rows or columns and provide a one dimensional result. Or 
it takes a one dimensional vector and returns a single value (zero dimensional vector).

@@

When compiling EmbArBB to ArBB a sharing detection pass is performed. This pass detects 
sharing in the AST using the method of A. Gill~\citet{Gill}. One benefit of this pass 
is that it reduces the numbers of calls into the ArBB C library. The efficiency of code 
generated is probably not affected since ArBB would discover that sharing in a Common 
Subexpression Elimination (CSE) pass. Another potential benefit is that sharing detection 
on the Haskell side results in a simpler AST being built on the ArBB side; potentially making 
the job for the ArBB optimisation passes simpler. What effects these 
operations truly had on performance has not been investigated. Actually, applying a sharing 
detection technique was one of my personal motivations for working on EmbArBB. In Obsidian 
we have not seen the need for this technique.

One benefit of a deep embedding is that transformations can be performed on the AST. Applying 
any optimisations in EmbArBB would very likely duplicate effort already put into ArBB. 
Therefore we apply no further transformations after sharing detection. However, if EmbArBB 
would be used to generate code in some other target language a set of optimisations 
would be needed.

@@

In paper H, section~\ref{sec:paperH}, we benchmark EmbArBB against ArBB in C++ and the Haskell 
Repa library. EmbArBB compares favourably to Repa and is often a lot faster. EmbArBB and the ArBB 
C++ performance is not distinguishable from EmbArBB's indicating that our Haskell embedding 
adds no extra overhead. Of course this is expected, since once the ArBB system has generated 
the code, that code is identical no matter if the AST was built via Haskell or C++ calls. 
Missing is a comparison of how well EmbArBB and C++ compares in regards to the time it actually 
takes to go through the process of making those ArBB C API calls. However, if the generated 
code is used many times the caching ArBB does of generated code means that the initial code 
generation cost can be amortised.

@@

In figure~\ref{fig:researchmatrix}, our work is placed in relation to other languages,
embedded languages and libraries for parallel programming. The languages and libraries considered 
are data-parallel; languages for control parallelism or distributed message passing systems are 
not considered. The systems are roughly divided 
into three groups, based on their level of abstraction. There are low-level languages; here 
I place languages that are imperative and C-like. In the middle layer I place languages 
that have higher level abstractions and are more easily composable. 
at the highest level, I place languages that completely abstract away from details of the 
hardware they run on. The division is not free from personal bias and the placement of languages
in boxes was not always easy.

@@

% ---------------------------------------------------------------------------
% Obsidian, Accelerate, Nikola, Thrust, CUB 
% ---------------------------------------------------------------------------
With Obsidian, we try to target the sparsely occupied area of mid-level GPU programming. With 
CUDA~\citet{wwwcuda} and OpenCL~\citet{OpenCL}, the programmer has full control of how to divide 
the computation amongst threads and blocks. Using Accelerate~\citet{ACCELERATEDAMP11}, 
Nikola~\citet{NIKOLA} 
and Thrust~\citet{THRUST} the programmer would use 
high-level patterns without direct insight into how the application would be mapped onto 
threads and blocks of the GPU. Using Obsidian, we want to bridge that gap by both being 
more composable than CUDA and giving the programmer control of what the computation will 
look like on the GPU. This is done by allowing the programmer to specify and compose 
thread-level and block-level code. Recently, the CUB library arrived that is built on a 
similar idea, that GPU programmers need to have block/thread level control to get maximum 
performance, while maintaining a higher level and more Thrust-like programmer 
interface~\citet{CUB}. Before CUB, and as far as I am aware, we were alone at targeting 
this level.

@@

% ---------------------------------------------------------------------------
% ArBB EmbArBB
% ---------------------------------------------------------------------------
The ArBB and EmbArBB systems also provide a set of parallel primitives that the programmer
composes to build his application. I place the ArBB and EmbArBB in a box slightly lower than 
Accelerate since while ArBB/EmbArBB also have built in reduction primitives, just like Accelerate, 
they are not general primitives. Where Accelerate has a single higher order reduction 
operation, ArBB/EmbArBB has \\ {\tt reduce\_add, reduce\_mul} and so on. Benefits of 
having ArBB embedded in Haskell compared to C++ are discussed in section~\ref{sec:EmbArBB}.
%I will go into 
%what benefits there are to having ArBB embedded in Haskell compared to C++ in

% ---------------------------------------------------------------------------
% Microsoft Accelerator
% ---------------------------------------------------------------------------
Microsoft Accelerator~\citet{ACCELERATOR}, is a language embedded in C\#. Accelerator has
data-parallel arrays and a set of aggregate operations (operations on entire arrays). The 
operations on data-parallel arrays are JIT-compiled to GPU code. Accelerator can also 
compile to multithreaded CPU code and make use of SIMD units of modern 
CPUs~\citet{ACCELERATORCPU}; this makes the Accelerator system related to ArBB.

% ---------------------------------------------------------------------------
% Python systems
% ---------------------------------------------------------------------------
There are many embedded languages implemented in Python that makes use of GPUs 
to improve performance. One example is Copperhead~\citet{copperhead}. Copperhead 
supports flat and nested data-parallelism via primitives such as {\tt map} and 
{\tt reduce}.

@@

Repa (Regular Shape-polymorphic Arrays)~\citet{REPA}, is a Haskell library for 
data-parallel programming. 
Repa uses the same array representation as Obsidian, in Repa called a delayed
array, which is a function from index to element. However, in Repa arrays have shape; 
they can be of any dimensionality and there are shape-polymorphic functions on these. 
For example, {\tt map} can be applied to an array of any shape.

Meta-Repa~\citet{METAREPA} is a reimplementation of the Repa library using deep embedded language 
techniques. Meta-Repa programs build ASTs that are then compiled (using template 
Haskell) to Haskell code using low-level parts of the Repa library for its parallelism.
The Meta-Repa approach to implementing the Repa library 
gives inlining of operations for free, where Repa programs need to be annotated 
with pragmas to ensure inlining. On the other hand, in Meta-Repa the programmer 
needs to explicitly state when something should not be inlined (using a {\tt force} primitive).
This is related to our work on EmbArBB that also uses an existing library (ArBB) 
for its parallelism. In reference~\citet{FPCDSL}, a case is made for this style 
of embedding preexisting DSLs.

@@

Delite~\citet{DELITE} is an infrastructure for implementation of domain 
specific embedded languages. There is an extensible internal representation (IR,AST); 
the programmer extends some base IR with domain specific constructs. The Delite system 
has been used to implement domain specific embedded languages for machine learning, rendering, 
physics and other domains. Delite, being a framework and the languages implemented using it 
very domains specific, was impossible to place in the matrix (figure~\ref{fig:researchmatrix}).

@@

Feldspar is an embedded language for digital signal processors~\citet{FELDSPAR}. Apart 
from using the syntactic library~\citet{SyntacticICFP12}, it is quite similar to Obsidian. 
There is a deeply embedded core language (called Core) and vectors as a shallow embedding 
on top of that, just like in Obsidian.

@@

In paper A~(section \ref{sec:paperA}) we describe our method for compiling monadic embedded 
languages. 
This means to take a monadic computation and observe its structure in order to generate
code. This work is closely related to a method described in ``The Constrained Monad Problem''
by Sculthorpe and Gill~\citet{sculthorpe2013constrained}. In that paper they solve a more 
general problem than just compilation, but our method is nice because it is easy to understand 
and solves a single problem and does it well. In ``Generic monad constructs'' Persson et al 
also solves the problem of compiling monads using a continuation monad.

@@

Obsidian is a language for GPU kernel implementation, not complete applications. 
It is not as well integrated into Haskell as for example Accelerate. There are 
ways to execute Obsidian kernels from within Haskell (we have seen examples of this in 
section~\ref{sec:GPUPROGRAMMING}); the recommended work flow, 
however, is to generate the CUDA as text and write the host code by hand. This 
can be seen as both a shortcoming and a feature. It is hard to integrate 
Accelerate programs in larger applications written in other languages; when 
using Obsidian a kernel is generated and can be used as part of for example a library.

Obsidian is also a quite low-level language. The programmer needs to know GPU details 
and is allowed to write programs at a level almost as low as CUDA, if it is desired. 
Paper E (section~\ref{sec:paperE}) contains examples of programming at this low level. 
One potential use of Obsidian could be as the code generating backend of high-level 
and very domain specific embedded languages. This could be a niche for Obsidian.

Warp level computations is work in progress and that these are not as well integrated 
into the hierarchy as the block, thread and grid levels are. The warp abstractions 
offered also requires more work in the code generating backend than the others, 
that correspond directly to CUDA concepts. The problem is that CUDA does not have 
a warp abstraction. Warps needs to be better integrated with the rest of Obsidian 
and the code generation required for warps needs more testing before being 
trustworthy and useful. When operational, warp level abstractions will help 
the programmer make better use of the GPU.

@@

This thesis contains two approaches to embedding a language for data-parallel programming 
in Haskell. EmbArBB embeds a preexisting library for data-parallelism in Haskell 
and at little effort from us, very good performance is gained. The performance of 
EmbArBB is entirely due to all the optimisation expertise and effort that was invested in 
the ArBB system. We get all that expertise for free. What EmbArBB provides is a 
way for Haskellers to benefit from the ArBB system. ArBB and EmbArBB are based on a 
set of parallel primitives for the programmer to use and compose. Obsidian takes 
a different approach. Built in primitives are very few and very low level, such as 
sequential and parallel for loops. On top of the low-level capabilities, a library 
of high-level primitives can be built. The key is that we want the programmer to 
be able to specify how to implement these higher level abstractions.

Looking at the related work, we see that almost all high level GPU programming 
languages provide abstractions of parallel primitives, aggregate operations, 
maps, folds and reductions, giving justification to our investigation of GPU 
programming without these. We believe that the GPU architecture is still volatile 
enough that there be no ``the right'' way of implementing these primitives. As of 
today, the programmer still needs to explore the design space and find a solution 
that performs well on a particular GPU.

Performance of Obsidian kernels is often good and the Obsidian code that they 
are generated from is often shorter than corresponding CUDA programs. Obsidian 
programs differ from CUDA programs by describing whole programs. With this I mean 
that a CUDA programmer needs to take a leap from the single threaded program 
parameterised on thread ID, to what actually will happen when this program is 
run by hundreds or thousands threads at once. The extensive indexing arithmetic 
often present in CUDA programs can make this leap a hard one. Similar programs 
in Obsidian could use operations on whole arrays such as {\tt map} or explicitly 
bounded parallel loops.

@@

\noindent This paper describes a simple and compositional method for 
reification of monads. This is useful for compilation of 
monadic embedded languages. We use a simple robot control language 
as the example. Robot programs can be expressed using Haskell do notation 
and is compiled down to a simple first order program representation. 
There is also a graphical simulator available that runs the compiled 
code. Robot language and simulator is available at github: \url{github.com/svenssonjoel/Robot}.

@@

The first real paper about Obsidian, our embedded language for GPU 
programming. In this paper we identify similarities between 
connection patterns in hardware and parallel computations on GPUs. 
We introduce a structured and compositional way to write data-parallel 
programs on a GPU.

@@

In this paper we describe a version of Obsidian implemented with a 
a different internal representation of GPU programs. We realised that 
most of our kernels have similar shape, they are parallel operations 
composed in sequence interspersed with barrier synchronisations. The 
internal representation used here captures this case exactly. Programming 
is now done is a style resembling arrows.

@@

This paper introduces {\em Push} arrays, invented by Koen Claessen. This 
array representation target specific performance problems that we want 
to solve with Obsidian. Here we have also gone back to using monadic 
style for programming in Obsidian.

@@

This paper contains a sorting case study performed in Obsidian. It 
is however the algorithms that are in focus here. We describe 
an interesting variation on Counting sort that removes duplicate 
elements and is a nice fit for GPUs because of its little need for 
synchronisation.

In order to implement these sorting algorithms we added atomic operations 
to Obsidian.

@@

In this paper the most recent additions and improvements of Obsidian are 
described. Obsidian Programs are now parameterised on Hierarchy level. This forces
the programmer into writing programs that we know how to compile to a GPU easily 
(by limiting how parallel operations can be nested). The paper also gives a detailed  
optimisation story using Obsidian for implementing reduction kernels.

>>>>

ArBB is based on a set of parallel primitives (or structures) on dense one, two and three dimensional vectors and nested vectors: 
\begin{itemize} 
\item {\bf Reductions:}   {\tt add\_reduce}, {\tt mul\_reduce}, {\tt max\_reduce} ... 
\item {\bf Scans:}        {\tt add\_scan}, {\tt mul\_scan}, {\tt max\_scan} ... 
\item {\bf Sorting:}      {\tt sort} 
\item {\bf Permutations:} {\tt gather}, {\tt scatter}, {\tt reverse} ...
\item {\bf Sequential loops:} {\tt \_for}, {\tt \_while} 
\item {\bf Map, zipWith, stencil:} are all supported via a single primitive called {\tt map}.
\end{itemize}

These operations (and many more, there are about 120 operations in total) can be used 
by the programmer when writing programs using ArBB. Note that there are sequential looping 
constructs with special names, {\tt \_for} and {\tt \_while}. ArBB is a deeply embedded 
language and calling the ArBB functions does not actually execute them, it just adds a node to 
some internal representation. Therefore, the use of normal C++ for loops leads to unrolled 
ArBB programs and a special {\tt \_for} loop is needed to get a loop in the generated program. 
This is exactly the way these things work in Haskell embedded languages, but it may be even 
more surprising to a C++ programmer. Before an ArBB function can be computed, it needs to be 
captured. For this, ArBB has a {\tt call} function that is used to run functions using ArBB 
functionality. The first time {\tt call} is used on a function pointer, that function is 
executed and the ArBB functions used in it build an AST (we say the function has been 
captured). The AST is then compiled and specialised for the hardware available. Any 
subsequent calls of a function using ArBB functionality do not lead to a recompilation; 
as the first call caches this compiled code.

@@

The ArBB Haskell bindings provide a very rudimentary interface to ArBB functionality and 
are not useful for application implementation. We need a higher-level interface to ArBB to offer 
to the Haskell programmer. Our first attempt was to implement an Accelerate backend 
using the ArBB bindings. However, this had some hard problems to solve. Accelerate is a 
richer language and there was an API mismatch between Accelerate and ArBB. These problems 
are mentioned in paper G, section~\ref{sec:paperG}. As a way to offer ArBB functionality 
to the Haskell programmer in a way that is useful and simpler to implement, we started 
working on EmbArBB.

EmbArBB improves the interfacing with Haskell by implementing a deeply embedded language. 
Now, programs in EmbArBB create ASTs. These ASTs are compiled via the ArBB bindings, creating 
function objects that can then be run.

@@

ArBB supports one, two and three dimensional arrays, called dense vectors. In \\ EmbArBB, we 
represent these with a data type, {\tt DVector}, parameterised by dimensionality and 
element type. The dimension parameter can be either {\tt Dim1}, {\tt Dim2} or {\tt Dim3}. 
There are also nested vectors, in EmbArBB called {\tt NVector}, parameterised on element type.

@@

EmbArBB provides many of the ArBB operations on vectors via a 
shape-polymorphic interface. For example, a reduction operation can take a two  
dimensional vector and reduce all rows or columns and provide a one dimensional result. 
It can also take a one dimensional vector and return a single value (zero dimensional vector).

@@

When compiling EmbArBB to ArBB, a sharing detection pass is performed. This pass detects 
sharing in the AST using the method of A. Gill~\citet{Gill}. One benefit of this pass 
is that it reduces the number of calls into the ArBB C library. The efficiency of code 
generated is probably not affected, since ArBB would discover that sharing in a Common 
Subexpression Elimination (CSE) pass. Another potential benefit is that sharing detection 
on the Haskell side results in a simpler AST being built on the ArBB side, potentially making 
the job for the ArBB optimisation passes simpler. What effects these 
operations truly had on performance has not been investigated. Actually, applying a sharing 
detection technique was one of my personal motivations for working on EmbArBB. In Obsidian, 
we have not seen the need for this technique. In Obsidian, sharing is in the hands of the 
programmer, via the {\tt force} operation. On a GPU, communication is more expensive than 
computation, so whether to share a computation or not becomes a tool for the programmer to use.

One benefit of a deep embedding is that transformations can be performed on the AST. Applying 
any optimisations in EmbArBB would very likely duplicate effort already put into ArBB. 
Therefore, we apply no further transformations after sharing detection. However, if EmbArBB 
would be used to generate code in some other target language a set of optimisations 
would be needed.

@@

In paper H, section~\ref{sec:paperH}, we benchmark EmbArBB against ArBB in C++ and the Haskell 
Repa library. EmbArBB compares favourably to Repa and is often a lot faster. The ArBB 
C++ performance is not distinguishable from EmbArBB's, indicating that our Haskell embedding 
adds no extra overhead. Of course, this is expected, since once the ArBB system has generated 
the code, that code is identical no matter if the AST was built via Haskell or C++ calls. 
A comparison of how well EmbArBB and C++ compares in regards to the time it actually 
takes to go through the process of making those ArBB C API calls has not been performed. 
However, if the generated code is used many times, ArBB's caching of generated code means 
that the initial code generation cost can be amortised.

@@

In figure~\ref{fig:researchmatrix}, our work is placed in relation to other languages,
embedded languages and libraries for parallel programming. The languages and libraries considered 
are data-parallel; languages for control parallelism or distributed message passing systems are 
not considered. The systems are roughly divided 
into three groups, based on their level of abstraction. There are low-level languages; here 
I place languages that are imperative and C-like. In the middle, layer I place languages 
that have higher level abstractions and are more easily composable. 
at the highest level, I place languages that completely abstract away from details of the 
hardware they run on. The division is not free from personal bias and the placement of languages
in boxes was not always easy.

@@

% ---------------------------------------------------------------------------
% Obsidian, Accelerate, Nikola, Thrust, CUB 
% ---------------------------------------------------------------------------
With Obsidian, we try to target the sparsely occupied area of mid-level GPU programming. With 
CUDA~\citet{wwwcuda} and OpenCL~\citet{OpenCL}, the programmer has full control of how to divide 
the computation amongst threads and blocks. Using Accelerate~\citet{ACCELERATEDAMP11}, 
Nikola~\citet{NIKOLA} 
and Thrust~\citet{THRUST}, the programmer uses 
high-level patterns, without direct insight into how the application is mapped onto 
the threads and blocks of the GPU. Using Obsidian, we want to bridge that gap by both being 
more composable than CUDA and giving the programmer control of what the computation will 
look like on the GPU. This is done by allowing the programmer to specify and compose 
thread-level and block-level code. Recently, the CUB library arrived that is built on a 
similar idea, that GPU programmers need to have block/thread level control to get maximum 
performance, while maintaining a higher level and more Thrust-like programmer 
interface~\citet{CUB}. Before CUB, and as far as I am aware, we were alone at targeting 
this level.

@@

% ---------------------------------------------------------------------------
% ArBB EmbArBB
% ---------------------------------------------------------------------------
The ArBB and EmbArBB systems also provide a set of parallel primitives that the programmer
composes to build his application. I place ArBB and EmbArBB in a box slightly lower than 
Accelerate,  since while ArBB/EmbArBB also have built in reduction primitives, just like 
Accelerate, 
they are not general primitives. Where Accelerate has a single higher order reduction 
operation, ArBB/EmbArBB has \\ {\tt reduce\_add, reduce\_mul} and so on. Benefits of 
having ArBB embedded in Haskell compared to C++ are discussed in section~\ref{sec:EmbArBB}.
%I will go into 
%what benefits there are to having ArBB embedded in Haskell compared to C++ in

% ---------------------------------------------------------------------------
% Microsoft Accelerator
% ---------------------------------------------------------------------------
Microsoft Accelerator~\citet{ACCELERATOR} is a language embedded in C\#. Accelerator has
data-parallel arrays and a set of aggregate operations (operations on entire arrays). The 
operations on data-parallel arrays are JIT-compiled to GPU code. Accelerator can also 
compile to multithreaded CPU code and make use of SIMD units of modern 
CPUs~\citet{ACCELERATORCPU}; this makes the Accelerator system related to ArBB.

% ---------------------------------------------------------------------------
% Python systems
% ---------------------------------------------------------------------------
There are many embedded languages implemented in Python that make use of GPUs 
to improve performance. One example is Copperhead~\citet{copperhead}. Copperhead 
supports flat and nested data-parallelism via primitives such as {\tt map} and 
{\tt reduce}.

@@

Repa (Regular Shape-polymorphic Arrays)~\citet{REPA} is a Haskell library for 
data-parallel programming. 
Repa uses the same array representation as Obsidian, in Repa called a delayed
array, which is a function from index to element. However, in Repa arrays have shape; 
they can be of any dimensionality and there are shape-polymorphic functions on these. 
For example, {\tt map} can be applied to an array of any shape.

Meta-Repa~\citet{METAREPA} is a reimplementation of the Repa library using deep embedded language 
techniques. Meta-Repa programs build ASTs that are then compiled (using template 
Haskell) to Haskell code that uses low-level parts of the Repa library for its parallelism.
The Meta-Repa approach to implementing the Repa library 
gives inlining of operations for free, where Repa programs need to be annotated 
with pragmas to ensure inlining. On the other hand, in Meta-Repa the programmer 
needs to explicitly state when something should not be inlined (using a {\tt force} primitive).
This is related to our work on EmbArBB that also uses an existing library (ArBB) 
for its parallelism. In reference~\citet{FPCDSL}, a case is made for this style 
of embedding preexisting DSLs.

@@

Delite~\citet{DELITE} is an infrastructure for implementation of domain 
specific embedded languages. There is an extensible internal representation (IR,AST); 
the programmer extends some base IR with domain specific constructs. The Delite system 
has been used to implement domain specific embedded languages for machine learning, rendering, 
physics and other domains. Delite, being a framework and the languages implemented using it 
very domains specific, was impossible to place in the matrix (Figure~\ref{fig:researchmatrix}).

@@

Feldspar is an embedded language for digital signal processing~\citet{FELDSPAR}. Apart 
from using the syntactic library~\citet{SyntacticICFP12}, it is quite similar to Obsidian. 
There is a deeply embedded core language (called Core) and vectors as a shallow embedding 
on top of that, just like in Obsidian.

@@

In paper A~(section \ref{sec:paperA}), we describe our method for compiling monadic embedded 
languages. This work is closely related to a method described in ``The Constrained Monad Problem''
by Sculthorpe and Gill~\citet{sculthorpe2013constrained}. In that paper, they solve a more 
general problem than just compilation, but our method is nice because it is easy to understand 
and solves a single problem and does it well. In ``Generic monad constructs'', Persson et al 
also solve the problem of compiling monads using a continuation monad.

@@

Obsidian is a language for GPU kernel implementation, not complete applications. 
It is not as well integrated into Haskell as for example Accelerate. There are 
ways to execute Obsidian kernels from within Haskell (we have seen an example of this in 
section~\ref{sec:GPUPROGRAMMING}); the recommended work flow, 
however, is to generate the CUDA as text and write the host code by hand. This 
can be seen as both a shortcoming and a feature. It is hard to integrate 
Accelerate programs in larger applications written in other languages; when 
using Obsidian a kernel is generated and can be used as part of for example a library.

Obsidian is also quite a low-level language. The programmer needs to know GPU details 
and is allowed to write programs at a level almost as low as CUDA, if it is desired. 
Paper E (section~\ref{sec:paperE}) contains examples of programming at this low level. 
One potential use of Obsidian could be as the code generating backend of high-level 
and very domain specific embedded languages. This could be a niche for Obsidian.

Obsidian's support for warp level computations is work in progress. Currently, warp level 
computations are not as well integrated into the hierarchy as the block, thread and grid 
levels are. The warp abstractions 
offered also require more work in the code generating backend than those 
that correspond directly to CUDA concepts. The problem is that CUDA does not have 
a warp abstraction. Warps needs to be better integrated with the rest of Obsidian 
and the code generation required for warps needs more testing before being 
trustworthy and useful. When operational, warp level abstractions will help 
the programmer make better use of the GPU.

@@

This thesis contains two approaches to embedding a language for data-parallel programming 
in Haskell. EmbArBB embeds a preexisting library for data-parallelism in Haskell 
and, at little effort from us, very good performance is gained. The performance of 
EmbArBB is entirely due to all the optimisation expertise and effort that was invested in 
the ArBB system. We get all that expertise for free. What EmbArBB provides is a 
way for Haskellers to benefit from the ArBB system. ArBB and EmbArBB are based on a 
set of parallel primitives for the programmer to use and compose. Obsidian takes 
a different approach. Built in primitives are very few and very low level, such as 
sequential and parallel for loops. On top of the low-level capabilities, a library 
of high-level primitives can be built. The key is that we want the programmer to 
be able to specify how to implement these higher level abstractions.

Looking at the related work, we see that almost all high level GPU programming 
languages provide abstractions of parallel primitives, aggregate operations, 
maps, folds and reductions, giving justification to our investigation of GPU 
programming without these. We believe that the GPU architecture is still volatile 
enough that there is no ``the right'' way of implementing these primitives. As of 
today, the programmer still needs to explore the design space and find a solution 
that performs well on a particular GPU.

Performance of Obsidian kernels is often good and the Obsidian code that they 
are generated from is often shorter than corresponding CUDA programs. Obsidian 
programs differ from CUDA programs by describing whole programs. A CUDA programmer 
needs to take a leap from the single threaded program 
parameterised on thread ID, to what actually will happen when this program is 
run by hundreds or thousands of threads at once. The extensive indexing arithmetic 
often present in CUDA programs can make this leap a hard one. Similar programs 
in Obsidian could use operations on whole arrays such as {\tt map} or explicitly 
bounded parallel loops.

@@

\noindent This paper describes a simple and compositional method for 
reification of monads. This is useful for compilation of 
monadic embedded languages. We use a simple robot control language 
as the example. Robot programs can be expressed using Haskell do notation 
and are compiled down to a simple first order program representation. 
There is also a graphical simulator available that runs the compiled 
code. The robot language and simulator are available at github: \\ 
\url{github.com/svenssonjoel/Robot}.

@@

This is the first real paper about Obsidian, our embedded language for GPU 
programming. In this paper, we identify similarities between 
connection patterns in hardware and parallel computations on GPUs. 
We introduce a structured and compositional way to write data-parallel 
programs on a GPU.

@@

In this paper we describe a version of Obsidian implemented with a 
a different internal representation of GPU programs. We realised that 
most of our kernels have similar shape; they are parallel operations 
composed in sequence interspersed with barrier synchronisations. The 
internal representation used here captures this case exactly. Programming 
is now done is a style resembling arrows.

@@

This paper introduces {\em Push} arrays, invented by Koen Claessen. This 
array representation targets specific performance problems that we want 
to solve with Obsidian. Here we have also gone back to using monadic 
style for programming in Obsidian.

@@

This paper contains a sorting case study performed in Obsidian. It 
is however the algorithms that are in focus here. We describe 
an interesting variation on Counting sort that removes duplicate 
elements and is a nice fit for GPUs because of its reduced need for 
synchronisation.

In order to implement these sorting algorithms, we added atomic operations 
to Obsidian.

@@

In this paper, the most recent additions and improvements to Obsidian are 
described. Obsidian Programs are now parameterised on Hierarchy level. This forces
the programmer into writing programs that we know how to compile to a GPU easily 
(by limiting how parallel operations can be nested). The central part of this paper 
describes how to implement and optimise reduction kernels using Obsidian.
